### 2018-04-08 Writing/disparity_review.tex
Removed definition box

One commonly used approach to analyse these discrete patterns and to understand the processes that led to it is to analyse the disparity of the morphological charactersitics of the group, often referred to as morphological disparity analysis.
This type of analysis was popularised around the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] by studying the fossil record and focusing on broad questions such how does diversity evolve, are some body types more common that others or can shape evolve in all ``directions'' (of the multidimensional space).
Since then, an impressive amount of literature and the equally vast amount of methods have been used and the term ``disparity'' has become a close synonym of ``analysis of patterns of morphological diversity'' and has been used in a range of sub-fields in biology.
Disparity analysis has been used to tackle a vast array of questions.
For example, it has been used (1) in Evo-devo to tackle questions of
%evo-devo
Developmental constraints [CITE];
Relationship between form and function and evolution functional traits [CITE] or
Scaling micro to macro [CITE];
% Ecology
(2) in ecology or palaeoecology to study:
intra-specific competition [CITE],
niche occupants, speciation and competition[CITE] or
macroecology (disparity in space)[CITE];
% Evolution
(3) on in macroevolution to study:
response to extinction and selectivity [CITE],
intra-specific competitive replacements [CITE] or
more globally, tempo and mode of evolution [CITE].

To date, however, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.
This diversity (and disparity) of methods to analysis disparity provides biologist with a powerful and comprehensive tool %TG: comprehensive in the sense "global"/"universal" not sure if that's the right word here.
that has a great deal of promise in helping us to understand the evolution of biological diversity.
However, as with any great power, it comes with great responsibility.

With the expansion of methodological approaches in the last decade, we also identify three main issues that, if not tackled could reduce the power of such analysis.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.

\subsubsection{[Box]Disparity patterns or processes}
\label{box_semantic}
At the core of defining what \textit{is} disparity lies a potential divide between epistemology and ontology: % NC: Can you either just write out what these terms mean, or add the explanation in brackets? I hate un-necessary jargon. Most PhD students will not be familiar with these terms. Were you before you got into this? (maybe you were, maybe the French education system is better!)
is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).
Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
Thus disparity % NC: I'm still unclear what you mean when you say disparity here. Do you mean the methods, the idea of discrete groups? I'm struggling a bit to parse how it can "inform us" per se.
could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
\subsubsection{[Box]End}
% NC: I think I prefer the previous version where you actually had a subsection to discuss this within the text rather than a box. Seems particularly weird as you als odo some defining of disparity in the first paragraphs. Maybe move that into the box? Then just say here we define dispairty as "XXXX" in the text. Or shorten this and put it within the text? See what works. But the point of a box is to add details not needed in the text, so if it's in the text anyway then I'm dubious about the box...

@@

>>>>

One commonly used approach to analyse these discrete patterns and to understand the processes that led to it is to analyse the disparity of the morphological charactersitics of the group, often referred to as morphological disparity analysis.
This type of analysis was popularised around the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] by studying the fossil record and focusing on broad questions such how does diversity evolve, are some body types more common that others or can shape evolve in all ``directions'' (of the multidimensional space).
Since then, an impressive amount of literature and the equally vast amount of methods have been used and the term ``disparity'' has become a close synonym of ``analysis of patterns of morphological diversity'' and has been used in a range of sub-fields in biology.
Disparity analysis has been used to tackle a vast array of questions in Evo-devo (Developmental constraints [CITE]; Relationship between form and function and evolution functional traits [CITE] or Scaling micro to macro [CITE];), in ecology or palaeoecology (intra-specific competition [CITE], niche occupants, speciation and competition[CITE] or macroecology (disparity in space)[CITE];)
or in macroevolution (response to extinction and selectivity [CITE], intra-specific competitive replacements [CITE] or more globally, tempo and mode of evolution [CITE]).

\subsection{What \textit{is} disparity?}

To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.

At the core of this debate a potential divide between whether disparity is a patterns or a processes: is there a difference between the disparity patterns (and how researchers characterise them) and how they were ultimately generated?
If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).

Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.
Thus, disparity methods or analysis can be seen as methods or analysis to study these patterns and eventually link them to macroevolutionary or macroecological processes

@@

This diversity (and disparity) of methods to analysis disparity provides biologist with a powerful and comprehensive tool %TG: comprehensive in the sense "global"/"universal" not sure if that's the right word here.
that has a great deal of promise in helping us to understand the evolution of biological diversity.
However, as with any great power, it comes with great responsibility. 
With the expansion of methodological approaches in the last decade, we also identify three main issues that, if not tackled could reduce the power of such analysis.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.
